# Chapter 1

## Section 1.2
Examples of time series data

### Example 1.1 J&J Quarterlies
```{r}
library(astsa)
plot(jj, type = "o", ylab = "Quarterly Earnings per Share")
```

Note the generally upward trend with minor variations. Variations
increase as time goes on, as well.

### Example 1.2 Global Warming
Time series data is its own data type in R with a default print function

```{r}
head(gtemp)
class(gtemp)
summary(gtemp)
str(gtemp)
plot(gtemp, type = "o", ylab = "Global Temperature Deviations")
```

Deviations from global mean temperature 1951 - 1980. Note the major
upward trend, but with other oscillations throughout.

### Example 1.3 Speech data
0.1 sec sample of recorded speech for *aaaaaaaaahhhhhh*.
```{r}
plot(speech)
```

Hugely repetitive, regular periodicities.  Can use spectral analysis
(eigenbased methods) to create "signature" in further analysis,
matching signature to other tokens can be used to identify
similarities, cluster, etc.

### Example 1.4 NYSE
```{r}
plot(nyse, ylab = "NYSE Returns")
```

Mostly stable average w/ lots of variance.  Huge drop/volatility
corresponds to drop in Oct. 1987.

#### Example 1.5 El Nino and Fish Pop.
Southern Oscillation Index (SOI) corresponds to El Nino, Recruitment
to number of new fish.

```{r}
par(mfrow = c(2, 1)) ## 2x1 grid for plotting
plot(soi, ylab = "", xlab = "", main = "Southern Oscillation Index")
plot(rec, ylab = "", xlab = "", main = "Recruitment")
par(mfrow = c(1, 1)) ## return to default
```

Both repetitive.  SOI is faster period than recruitment.  Recruitment
has two underlying oscillations. Faster period repeating yearly, but
notice drops approx. 4 years apart.

### Example 1.6 fMRI Imaging
fMRI data (brain imaging) from various brain locations on five
subjects w/ periodic brushing of their hands.  Response measures
levels of activation in brain.

```{r}
par(
    mfrow = c(2, 1), # stack images
    mar = c(3, 2, 1, 0) + 0.5, # set margins
    mgp = c(1.6, .6, 0)) # margin line (closer titles)
```

### Example 1.2 Global Warming
Time series data is its own data type in R with a default print function

```{r}
head(gtemp)
class(gtemp)
summary(gtemp)
str(gtemp)
plot(gtemp, type = "o", ylab = "Global Temperature Deviations")
```

Deviations from global mean temperature 1951 - 1980. Note the major
upward trend, but with other oscillations throughout.

### Example 1.3 Speech data
0.1 sec sample of recorded speech for *aaaaaaaaahhhhhh*.
```{r}
plot(speech)
```

Hugely repetitive, regular periodicities.  Can use spectral analysis
(eigenbased methods) to create "signature" in further analysis,
matching signature to other tokens can be used to identify
similarities, cluster, etc.

### Example 1.4 NYSE
```{r}
plot(nyse, ylab = "NYSE Returns")
```

Mostly stable average w/ lots of variance.  Huge drop/volatility
corresponds to drop in Oct. 1987.

### Example 1.5 El Nino and Fish Pop.
Southern Oscillation Index (SOI) corresponds to El Nino, Recruitment
to number of new fish.

```{r}
par(mfrow = c(2, 1)) ## 2x1 grid for plotting
plot(soi, ylab = "", xlab = "", main = "Southern Oscillation Index")
plot(rec, ylab = "", xlab = "", main = "Recruitment")
par(mfrow = c(1, 1)) ## return to default
```

Both repetitive.  SOI is faster period than recruitment.  Recruitment
has two underlying oscillations. Faster period repeating yearly, but
notice drops approx. 4 years apart.

### Example 1.6 fMRI Imaging
fMRI data (brain imaging) from various brain locations on five
subjects w/ periodic brushing of their hands.  Response measures
levels of activation in brain.

```{r}
par(mfrow = c(2, 1), # stack images
    mar = c(3, 2, 1, 0) + 0.5, # set margins
    mgp = c(1.6, .6, 0)) # margin line (closer titles)
ts.plot(fmri1[,2:5],
        lty = c(1:2, 4:5),
        ylab = "BOLD",
        xlab = "",
        main = "Cortex")
ts.plot(fmri1[,6:9],
        lty = c(1:2, 4:5),
        ylab = "BOLD",
        xlab = "",
        main = "Thalamus & Cerebellum")
```

Similar periods, but T&C has more variance & lower amplitude than
Cortex. Both respond to the stimulus, but cortex responds more strongly.

### Example 1.7 Earthquakes & Explosions
Seismic data from the arrivals of waves from a recording station in
Scandinavia.

```{r}
par(mfrow = c(2, 1))
plot(EQ5, main = "Earthquake")
plot(EXP6, main = "Explosion")
```

Explosions start with a small burst, become small waves, then rapidly
increase in amplitude before steadily diminishing.

Quakes start small, increase to a steady but moderately strong wave,
then increase to large amplitude, but with a longer period. Waves
decrease in strength slightly, then stay
generally the same amplitude until the end of recording.

## Section 1.3 Time Series Statistical Models
 - Need math/stat models that give plausible descriptions for time and
   frequency dependent data.
 - Assume TS data is defined as a collection of RVs indexed according
   to time they are observed.
 - Denotes $\{x_t\}$ as vector of random variables indexed by $t$
   s.t. $t = 0, \pm 1, \pm 2, \ldots$ or some subset of integers and
   values are observed for each time period, in order, and refer to
   a stochastic (random) process.
 - The observed values of $\{x_t\}$ are a realization of the
   stochastic process, and referred to as the time series.
 - Generally we *can* observe the values at any continuous point in time
   and conceptually are dealing with continuous RVs. Because of
   restrictions in data collection, we are typically forced to use a
   discrete time parameter, however.
 - Theoretical developments assume that continuous parameter time
   series should be specified in terms of finite-dimensional
   distribution functions defined over a *finite* number of points,
   allowing for a suitable definition of the periods indexed by $t$.
 - Insufficient sampling rates can lead to aliasing, discussed later.
 - Generally we suppose that adjacent points in time are correlated,
   so $x_t$ depends in some way on the values $x_{t - 1}, x_{t - 2},
   \ldots$

## Example 1.8 White Noise
say we have uncorrelated random variables, $w_t$ with mean 0 and
variance $\sigma^2_w$. The TS generated from uncorrelated variables is
*white noise*. Denote white noise processes as $w_t \sim wn(0,
\sigma^2_w)$. This is analogous to white light, indicating that all
possible period oscillations (waves) are present with equal strength.

When necessary, iid assumptions for $w_t$ are represented as $w_t \sim
iid(0, \sigma^2_w)$

If the stochastic behavior could simply be explained by white noise,
we can use classical statistics because of the iid
assumption. However, serial correlation and smoothness of TS data
requires more sophisticated methods.

### Example 1.9 Moving Averages
Instead of a white noise series $w_t$, we can replace it with a moving
average that smooths the series. For example, replace the noise with
an average of the current value and its immediate neighbors (both past
& future).

$$ v_t = \frac{1}{3}\sum_{t = -1}^{1} w_t$$

```{r}
w <- rnorm(500, 0, 1)
v <- filter(w, sides = 2, filter = rep(1/3, 3))
par(mfrow = c(2, 1))
plot.ts(w, main = "White Noise")
plot.ts(v, main = "Moving Average")
```

A linear combination of values in a TS are referred to as a
filtered series, hence `filter`.

Note that the moving average *smooths* the TS, making slower
oscillations more obvious but removing or diminishing more rapid
oscillations. See the SOI and fMRI figures above.

The speech series and Recruitment series differ from the moving
average because one type of pattern in particular is predominant,
generating a sinusoidal wave.

## Example 1.10 Autoregressions
Suppose $w_t$ is simply an input to a second-order equation:

$$x_t = x_{t-1} - 0.9x_{t-2} + w_t$$

for $t = 1, 2, \ldots, 500$. In this case, the equation represents a
regression of the current value of $x_t$ of the TS as a function of
the past two values of the series, leading us to the term
*autoregression*. Note that this leads to a problem with startup
values, because the equation also depends on the initial conditions
$x_0$ and $x_{-1}$. For now, assume we know these values.

```{r}
w <- rnorm(550, 0, 1) #extra 50 to avoid startup issues
x <- filter(w, filter = c(1, -0.9), method = "recursive")[-(1:50)]
plot.ts(x, main = "autogression")
```

Note the periodic behavior, similar to the speech data above.

Autoregressive models are widely used for many observed TSs.

## Example 1.11 Random Walk with Drift
For trends like the global warming data, we often use a random walk
with drift:

$$x_t = \delta + x_{t-1}+w_t$$

with initial condition $x_0 = 0$, where $w_t$ is white noise. $\delta$
represents the drift, and $\delta = 0$ denotes a random walk. We
describe it as such because the value at time $t$ is the value at time
$t-1$ plus some random movement determined by $w_t$. This model can
also be written as an expression of the cumulative sum of the white
noise:

$$x_t = \delta t + \sum_{j = 1}^t w_j$$

```{r}
set.seed(154)
## without drift
w <- rnorm(200, 0, 1)
x <- cumsum(w)
## with drift
wd <- w + .2
xd <- cumsum(wd)
plot.ts(xd, ylim = c(-5, 55), main = "random walk")
lines(x, lty = "dotted") ## no drift
lines(.2*(1:200), lty = "dashed") ## y = 0 + 0.2x
```

